---
title: "Machine Learning 1"
author: "Nicole Manuguid"
date: "10/21/2021"
output: pdf_document
---

First up is clustering methods 

# Kmeans clustering 

THe function in base R to do Kmeans clustering is called 'kmeans()'

First make up some data where we know what the answer should be: 

rnorm gives you 30 values around -/+3
```{r}
tmp <- c(rnorm(30,-3), rnorm(30, 3))
x <- cbind( x=tmp, y=rev(tmp))
x
plot(x)
```

> Q. Can we use kmeans() to cluster this data setting k to 2 and nstart to 20?

```{r}
km <- kmeans(x, centers = 2, nstart = 20)
km
```

> Q. How many points are in each cluster?

```{r}
km$size
```

> Q. What 'component' of your result object details cluster assignmnet/membership?

```{r}
km$clusters
```


>Q. What 'component' of your result object details clsuter center?

```{r}
km$centers

```


>Q. Plot x colored by the kmeans cluster assignment and add cluster centers as blue points

```{r}
plot(x, col=km$cluster)
points(km$centers, col = "blue", pch = 15, cex = 2)
```


```{r}
tmp <- c(rnorm(30,-3), rnorm(30, 3))
x <- cbind( x=tmp, y=rev(tmp))
x
plot(x)
```


# Hierarachical Clustering

A big limitation with kmeans is that we have to tell it K (the number of clusters we want)

Analyze this same data with hclust()

Demonstrate the use of dist(), hclust(), plot (), and cutree() functions to do clustering,
Generate dendrograms and return cluster assingment membership vector...

```{r}
hc <- hclust(dist(x))
hc
```

There is a plot method for hclust result objects. Let's see it

```{r}
plot(hc)
```
To get our cluster membership vector we have to do a wee bit more work.We have to "cut" the tree where we think it makes sense. For this we use the 'cutree()' function 

```{r}
cutree(hc, h = 6)
```

You can also call 'cutree()' setting k=the number of grps/clusters you want. 

```{r}
grps <- cutree(hc, k = 2)
```


Make our results plot

```{r}
plot(x, col=grps)
```


#Principal Component Analysis(PCA) of UK Food Data

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)

```

Checking our data

```{r}
View(x)
```

fix row names; this should be 17 x 4 but we have 17 x5 dimensions 

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)

```
recheck dimensions

```{r}
dim(x)
```
> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

be careful with previous function. can use this function below to get the same results


```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names = 1)
head(x)
```

Spotting major differences and trends

```{r}
barplot(as.matrix(x), beside = T,  col=rainbow(nrow(x)))
```

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

```{r}
barplot(as.matrix(x), beside = FALSE,  col=rainbow(nrow(x)))
```

> Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(10), pch = 16)
```
If a point lies in the diagonal between two countries it means they share similarities within the data. 

> Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

For N.Ireland, the blue point is not following the general shape of the diagonal and is most different compared to other coutnries in the UK.

#PCA to the rescue

the main function is in base R

the main function in base R for PCA is 'prcomp()' 
this wants the transpose of our data

```{r}
t(x)
```


```{r}
pca <- prcomp(t(x))
summary (pca)
```


> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim =c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))
     
```

Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim =c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col = c("yellow", "red", "blue", "green"))

```


calculate how much variation in the original data each PC accounts for 

```{r}
v <- round(pca$sdev^2/sum(pca$sdev^2)*100)
v
```

```{r}
z <- summary(pca)
z$importance
```


```{r}
barplot(v, xlab = "Principal Component", ylab = "Percent Variation")
```

```{r}
par(mar=c(10, 3,0.35, 0))
barplot (pca$rotation[,1], las=2)
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
par(mar=c(10, 3,0.35, 0))
barplot (pca$rotation[,2], las=2)
```

The two main food groups are fresh potatoes and soft drinks. PC2 shows us how much variation there is in food in N.Ireland compared to the rest of the countries in the UK. 


#biplots

```{r}
biplot(pca)
```

#pca of r-seq data 

```{r}
url2 <-"https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)


```

 > Q10: How many genes and samples are in this data set?

```{r}
dim(rna.data)
```

There are 100 genes and 10 samples 

```{r}
pca <- prcomp(t(rna.data), scale = TRUE)
plot(pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2")
```
```{r}
summary(pca)
```

```{r}
plot (pca, main = "Quick scree plot")
```

## Variance captured per PC 
```{r}
pca.var <- pca$sdev^2

```

## Percent variance is more informative to look at 

```{r}
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

generate scree-plot

```{r}
barplot(pca.var.per, main="Scree plot", name.arg = paste0 ("PC", 1:10), xlab = "Principal Component", ylab = "Percent Variation")

```

## A vector of colors for wt and ko samples


```{r}
colvec <- colnames(rna.data)
colvec[grep ("wt", colvec)] <- "red"
colvec[grep ("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col = colvec, pch = 16, xlab = paste0("PC1 (", pca.var.per[1], "%"), ylab = paste0("PC2 (", pca.var.per[2], "%"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
text
```



```{r}
library(ggplot2)
df <- as.data.frame(pca$x)

#make first plot

ggplot(df) + aes(PC1, PC2) + geom_point()
```


```{r}
df$samples <- colnames(rna.data)
df$condition <- substr(colnames(rna.data), 1,2)

p <- ggplot(df) + aes(PC1, PC2, label = samples, col = condition) + geom_label(show.legend = FALSE)

p
```


polish plot

```{r}
p + labs(title = "PCA of RNASeq Data", subtitle = "PC1 clearly separates wild-type from knock-out samples", x = paste0("PC1 (", pca.var.per[1], "%)"), y=paste0("PC2 (", pca.var.per[2], "%)"), caption = "BIMM143 example data") + theme_bw()

```

