---
title: "class 17"
author: "Nicole Manuguid"
date: "11/23/2021"
output:
  pdf_document: 
    latex_engine: xelatex
---

#Covid Vaccination Rates

We will take data from the CA.gov site here:

Statewide COVID-19 Vaccines Administered by ZIP Code” CSV file from: https://data.ca.gov/dataset/covid-19-vaccine-progress-dashboard-data-by-zip-code

```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

>Q1. What column details the total number of people fully vaccinated? 

persons fully vaccinated

>Q2. What column details the Zip code tabulation area? 

zip code tabulation data 

>Q3. What is the earliest date in this dataset? 

21-01-05

>Q4. What is the latest date in this dataset?

21-11-16

## Quick look at data structure 

As before we can use the **skimr()** function to quicky overview and summarize the dataset
```{r}
library(skimr)
```

```{r}
skimr::skim(vax)
```

>Q5. How many numeric columns are in this dataset? 

9

>Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 

```{r}
sum( is.na(vax$persons_fully_vaccinated) )
```

>Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 

```{r}
round (sum( is.na(vax$persons_fully_vaccinated) )/nrow(vax)*100,2)
```

#Ensure the data column is useful

We will use the **lubridate** package to make life a lot easier 

```{r}

#install.packages("lubridate")

```

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
# Speciffy that we are using the Year-mont-day format
vax$as_of_date <- ymd(vax$as_of_date)
```

> Q. how many days since the first entry?

```{r}
today() - vax$as_of_date[1]
```

> >Q9. How many days have passed since the last update of the dataset?

```{r}
vax$as_of_date[nrow(vax)]-vax$as_of_date[1]
```


>Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 

```{r}
length(unique(vax$as_of_date))
```


## Working with ZIP codes

We will use the **zipcodeR** package to help make sense of zip codes 
```{r}
#install.packages("zipcodeR")
```

```{r}
library(zipcodeR)
```

```{r}
geocode_zip('92037')

```

```{r}
zip_distance('92037','92109')
```
```{r}
reverse_zipcode(c('92037', "92109") )
```

#Foucs on San Diego County

```{r}
library(dplyr)
```

```{r}
table(vax$county)
```

We can subset with base R
```{r}
inds <- vax$county == "San Diego"
head(vax[inds,])
```
Use **dplyr** package and it's **filter** function: 
```{r}
sd <- filter(vax, county == "San Diego")

#How many entries are there for San Diego county?
nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
#sd.10

```

>Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```

>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
ind <-which.max(sd$age12_plus_population)
sd[ind,]
```

What is the population in the 92037 ZIP code area?
```{r}
filter(sd, zip_code_tabulation_area =="92037")[1,]
```

>Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2021-11-09”? 

```{r}
sd.now <- filter(sd, as_of_date == "2021-11-09")

mean(sd.now$percent_of_population_fully_vaccinated, na.rm = TRUE)
```
We can look at the 6-number summary

```{r}
summary(sd.now$percent_of_population_fully_vaccinated)
```

>Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2021-11-09”?

Using base R plots

```{r}
hist(sd.now$percent_of_population_fully_vaccinated)
```

```{r}
library(ggplot2)
```

```{r}
ggplot(sd.now) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram(bins = 15)
```

#Focus on UCSD/La Jolla

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

>Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point()+
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
  
```

#Comparing 92037 to other similar sized areas?

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

>Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
ucsd.now <- filter(vax.36, as_of_date == "2021-11-16")
mean(ucsd.now$percent_of_population_fully_vaccinated, na.rm = TRUE)
```
#Time series of vaccination rate for 92037

First select all data for the UCSD 92037 area

```{r}
ucsd <- filter(vax, zip_code_tabulation_area == "92037")
```

```{r}
ggplot(ucsd) +
  aes(as_of_date,percent_of_population_fully_vaccinated) +
  geom_point()+
  geom_line(group=1) +
  geom_hline(yintercept = 0.6629812, col="red", linetype="dashed")+
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated")
```

Population in the 92037 ZIP code area
```{r}
ucsd[1,]$age5_plus_population
```
First we need to subset the full 'vax' dataset to include only ZIP code areas iwth a population as large as 92037

```{r}
vax.36.all <-filter(vax, age5_plus_population > 36144)
nrow(vax.36.all)
```

How many unique zip codes have a population as large as 92037?
```{r}
length(unique(vax.36.all$zip_code_tabulation_area))
```

>Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”?

```{r}
summary(ucsd.now$percent_of_population_fully_vaccinated)
```


>Q18. Using ggplot generate a histogram of this data.

```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")

head(vax.36)
```

```{r}
ggplot(vax.36) + 
  aes(percent_of_population_fully_vaccinated) +
  geom_histogram()
```

>Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2021-11-16") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

>Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144

```{r}
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated,
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2) +
  geom_hline(yintercept = 0.66, col="red", linetype="dashed")
  
```

>Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?

I am unsure how I feel given the data and observations. Living in California I always thought there was a higher amount of people vaccinated but the data says otherwise. I think its inevitable that people will meet and gather so I'm feeling the same about in-person class, all I can do is try to protect myself by stayin safe.